---
title: "Python with reticulate"
teaching: 15
exercises: 5
---

```{r python-setup, include=FALSE}
# Ignore any conda/system Python on the PATH and use an environment that
# reticulate builds itself. Without this, a miniforge install earlier in the
# order of discovery wins and the py_require() declarations below are ignored.
Sys.setenv(RETICULATE_PYTHON = "managed")

library(reticulate)

py_require(
  packages       = c("pandas", "matplotlib"),
  python_version = "3.12",
  # Cap dependency resolution at a date. renv.lock records R packages only and
  # will never pin Python ones, so this is what keeps the episode reproducible.
  exclude_newer  = "2026-08-01"
)
```

:::::::::::::::::::::::::::::::::::::: questions

- How do I run Python code in a Workbench lesson?
- How do I declare the Python packages an episode needs?
- How do I pass objects between R and Python?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Declare an episode's Python dependencies with `py_require()`
- Write and execute `{python}` chunks alongside `{r}` chunks
- Move data across the language boundary with `py$` and `r.`

::::::::::::::::::::::::::::::::::::::::::::::::

## Declaring what the episode needs

Delete this episode if your lesson is R only. If it is not, everything you need is in the hidden
`python-setup` chunk at the top of this file, which reads:

```r
Sys.setenv(RETICULATE_PYTHON = "managed")
library(reticulate)
py_require(
  packages       = c("pandas", "matplotlib"),
  python_version = "3.12",
  exclude_newer  = "2026-08-01"
)
```

Two conventions to follow when you copy it. Give the chunk `include=FALSE` so the plumbing does
not appear in the rendered episode, and **do not label it `setup`**: sandpaper injects a chunk of
that name into every episode and knitr rejects duplicate labels.

`py_require()` does not install anything on its own. It records what the session needs, and the
first time a `{python}` chunk runs, `reticulate` uses [uv](https://docs.astral.sh/uv/) to build an
ephemeral virtual environment satisfying the whole set, downloading the interpreter as well as
the packages. Nothing lands in your system Python and nothing needs to exist beforehand.

Version constraints use pip syntax, so `"pandas>=2.2"` and `"scikit-learn==1.5.2"` both work.

::::::::::::::::::::::::::::::::::::: callout

## Why `RETICULATE_PYTHON = "managed"`

`reticulate` walks an order of discovery and only falls back to an ephemeral environment if it
finds no Python earlier in that order. On a machine with conda or miniforge on the `PATH`, conda
wins, your `py_require()` calls are silently ignored, and the episode renders against whatever
happens to be in that environment. It then fails in CI, where that environment does not exist.
Setting `"managed"` forces the reproducible path in both places.

::::::::::::::::::::::::::::::::::::::::::::::::

## Running Python

A `{python}` chunk works exactly like an `{r}` chunk:

```{python}
import pandas as pd

measurements = pd.DataFrame({
    "sample":  ["A", "B", "C", "D"],
    "depth":   [10, 25, 40, 55],
    "reads":   [1_204_000, 987_000, 1_530_000, 742_000],
})

measurements
```

Objects persist between Python chunks in the same episode, in one shared session:

```{python}
measurements["reads_per_m"] = measurements["reads"] / 1e6
measurements.describe()
```

## Crossing the language boundary

Python objects are reachable from R through `py$`, converted to their R equivalents. A pandas
`DataFrame` arrives as a `data.frame`:

```{r}
str(py$measurements)
summary(py$measurements$reads)
```

R objects go the other way through `r.`:

```{r}
threshold <- 1e6
```

```{python}
above = measurements[measurements["reads"] > r.threshold]
above[["sample", "reads"]]
```

## Plots

Matplotlib figures render like any other chunk output. Set `fig.alt` for accessibility, exactly
as you would for an R plot:

```{python depth-vs-reads, fig.alt = "Scatter plot of sequencing depth against read count for four samples"}
import matplotlib.pyplot as plt

fig, ax = plt.subplots(figsize = (5, 3.5))
ax.scatter(measurements["depth"], measurements["reads_per_m"])
ax.set_xlabel("Depth")
ax.set_ylabel("Reads (millions)")
fig.tight_layout()
plt.show()
```

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 1: add a dependency

The episode needs `numpy` as well. What do you change, and what do you commit?

:::::::::::::::::::::::: solution

Add it to the `py_require()` call in the setup chunk:

```r
py_require(c("pandas", "matplotlib", "numpy"), python_version = "3.12")
```

Nothing else. Python packages are not recorded in `renv.lock`, so the `.Rmd` file itself is the
only thing to commit. Contrast this with adding an **R** package, where you add `library(...)`,
rebuild, and then also commit the changed `renv/profiles/lesson-requirements/renv.lock`.

:::::::::::::::::::::::::::::::::

## Challenge 2: which Python is actually running?

How would you check, from inside the episode, that the managed environment is being used rather
than a conda install?

:::::::::::::::::::::::: solution

```r
reticulate::py_config()
```

The reported `python` path should sit under reticulate's own cache directory, not under
`miniforge3/` or `/usr/bin`.

:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

## When this is not enough

`py_require()` resolves from PyPI. If your lesson needs a bioconda stack, or tools that are not
packaged for PyPI, commit an `environment.yml`, point `RETICULATE_PYTHON` at that environment's
interpreter instead of `"managed"`, and add a conda setup step to
`.github/workflows/docker_build_deploy.yaml`. That is considerably more to maintain, so check
first whether the episode should be *showing* pipeline output rather than executing a pipeline at
build time.

::::::::::::::::::::::::::::::::::::: keypoints

- `Sys.setenv(RETICULATE_PYTHON = "managed")` forces the reproducible environment; without it a
  local conda install wins and the build differs from CI
- `py_require()` declares packages and the Python version; `uv` provisions them on first use
- `exclude_newer` caps resolution at a date, since `renv.lock` never records Python packages
- `py$name` reads Python objects from R, `r.name` reads R objects from Python
- No CI change is needed: `uv` fetches the interpreter inside the build container

::::::::::::::::::::::::::::::::::::::::::::::::
